---
title: "Connect R to a database"
author: "Matthew Van Scoyoc"
date: "August 30, 2016"
output: html_document
---

```{r setup, eval=TRUE, echo=TRUE, message=FALSE}
#' Packages
#install.packages(c("RODBC", "dplyr", "tidyr", "knitr", "ggplot2"))
library("RODBC")
library("dplyr")
library("tidyr")
library("knitr")
library("ggplot2")
```

## Connect database

Here we will use the RODBC package to connect a Microsoft Access 2010 database 
to R and read a table into R as a dataframe. Let't start by connecting the 
database to R by using the odbcConnectAccess2007() command.

```{r data, eval=TRUE, echo=TRUE}
#' Open database connections
my.dir <- "P:/1_ResourceStewardshipScience/Natural Resource Program/1 CARE range assessment project/4 Data & Photos/CARE Range DB"
my.db <- "CARE_RangelandAssessmentData_Master.accdb"
data.db <- odbcConnectAccess2007(paste(my.dir, my.db, sep = "/"))
```

## Read in tables

Now let's use the sqlFetch() command to read a couple tables into R as 
dataframes. We'll start with the tblPlots_Sampled table and produce a quick 
summary of the number of plots sampled per allotment.

```{r plots.sampd, eval=TRUE, echo=TRUE}
#' Let's look at the number of plots sampled per allotment
plots.sampd <- sqlFetch(data.db, "tblPlots_Sampled")
kable(plots.sampd %>% 
        group_by(Allotment) %>%
        summarise(n())
      )
```

Now lets read in the tblPellet_Counts table and summarize it for a figure.

```{r scat, eval=TRUE, echo=TRUE, message=FALSE}
#' Read in and subset scat data
scat.cow.v.elk <- tbl_df(sqlFetch(data.db, "tblPellet_Counts")) %>%
  filter(Animal == "cattle" | Animal == "elk") %>%
  group_by(Plot_ID, Animal) %>%
  summarise(Total = sum(Tally)) %>%
  left_join(select(plots.sampd, Plot_ID, Allotment)) %>%
  group_by(Allotment, Animal) %>%
  summarise(Total = sum(Total))

kable(scat.cow.v.elk)
```

## Figures

Lastly, let's see what the scat data look like.

```{r pressure, echo=TRUE}
#' Make a bar plot of cattle and elk scat observed during the assessments.
g <- ggplot(scat.cow.v.elk, aes(x = Animal, y = Total, fill = Animal)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Allotment) +
  theme_bw()
print(g)
```

## Close database

Okay, that was fun. Now we can close the database and remove it from the 
workspace.

```{r close, echo=TRUE}
#+ Close DB, eval = TRUE, echo = TRUE ----
odbcClose(data.db)
rm(data.db)

#+ EndScript ----
Sys.Date()
sessionInfo()
```
